### 2010-03-31 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: fix replace, adding hindx case
Desc@ICFP: proofread Section 4 'Levitating Desc'. Next ICFP: in Esperanto.

Apart from these changes, we are left with implementing the other
components of finite sets. This consists in $\EnumT$, $\Ze$, $\Su$, as
well as the $\F{switch}$ operator. Note that the actual implementation
of $\EnumU$ does not influence our implementation of $\EnumT$: be they
hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects
behave the same. We have witnessed this effect when carrying the
operation in Epigram, moving from an hard-coded presentation to a
self-hosted one. Absolutely no change to the $\EnumT$ objects was
required.

\begin{wstructure}
<- Summary of the operation
    <- The content of the type theory is exactly the same
        <- before == after
    /> type naming scheme condenses
        <- Replace named constructors by codes in the universe of data-types
    -> Our next step is a similar move (in essence)
        /> Condenses the entire naming scheme of data-types
\end{wstructure}

@@

If not on the content, this modification had an effect on the
names. The type naming scheme of the type theory has condensed: named
type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
now replaced by codes and their fix-point in the universe of
descriptions. In essence, our next step is similar: we are going to
condense the entire naming scheme of data-types \emph{in itself}.

@@

\begin{wstructure}
<- Realising our promises
    <- We are going to implement Desc
    /> Desc is itself a data-type
        <- Same situation as EnumU
            <- We want to benefit from generic operations
        -> It ought to be encoded in itself
\end{wstructure}

We shall now fulfil our promises for $\Desc$. We are going to
implement the codes of the universe of descriptions. In and by itself,
the system of codes, $\Desc$, is nothing but a data-type. We are in
the same situation as with $\EnumU$: we ought to be able to describe
the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
first-class citizen, born with the standard, generic equipment of
data-types.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's constructors:
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
that the tagged notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are
defining.

The case of $\DUnit$ is trivial: we immediately reach the end of the
data-type. $\DUnit$ has no payload. The case of $\DIndx{\!}$ comes
naturally: it only takes a recursive argument -- a description. The
case of $\DSigma{\!}{\!}$ is problematic. Recall the specification of
$\DSigma{\!}{\!}$:

@@

Equipped with $\DHindx{\!}{\!}$, we can describe the extended code of
our universe of data-types, as shown in
Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
remain unchanged, as expected. We successfully describe the
$\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing the $\V{H}$
in $\Set$ with a recursive argument.

At a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fix-point of this
object gives us $\Desc$, up to isomorphism. We have implemented our
code of data-types as an object levitating inside itself. However,
this levitation operation, just as any magic trick, relies on an
invisible cable. Let us reveal it, hence finishing our implementation.

@@

\begin{wstructure}
<- Generic programming now!
    <- Desc is just data
        -> Can be manipulated
    <- Free induction scheme on Desc
        -> Ability to inspect data-types
        -> Ability to program on data-types
\end{wstructure}

@@

This concludes our levitation work. Beyond its pedagogical value, this
exercise has several practical outcomes. First of all, this exercise
reveals that the $\Desc$ universe is just plain data. Just as any
piece of data, it can therefore be inspected and
manipulated. Moreover, it is expressed in the $\Desc$ universe. As a
consequence, it is equipped, for free, with an induction
principle. So, our ability to inspect and program with $\Desc$ is not
restricted to the meta-language: we now have all the necessary
equipment in the target language to \emph{program} over data-types. In
this setting, \emph{generic programming is just about programming}.

@@

\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
    \Void                                                                                                         \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
    \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto       \\
\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                     \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{xs}}{}\: \pair{\V{g}}{\V{ys}}{} \mapsto   \\
\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\{ \V{X}\: \V{Y}\: \V{g}\: \V{ys}}{}
}\]

Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\: (\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the
problem. In the type theory, we have built a generic catamorphism. Any
data-type will now come equipped with this operation, for free.

@@

\begin{wstructure}
<- Deriving generic functions
    <- Taking a Desc and computing a function
        <- Desc comes equipped with an induction principle
        -> Ability to compute more functions from it
            -> More generic functions
    <- Inspecting data-types
        <- All described byu a Desc code
        -> Ability to explore the code
            <- Desc equipped with an induction principle
            -> Build new objects based on that structure
\end{wstructure}

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a data-type specific
operation. This programming style is reminiscent to polytypic
languages, such as PolyP~\cite{jansson:polyp}.

Moreover, the $\F{replace}$ function demonstrates the benefit of an
approach based on universes. The data-types living in the universe of
descriptions, we are able to \emph{inspect} them. As shown by
$\F{replace}$, it is easy to explore these structures, as well as
building new ones.

@@

In this section, we will turn to a more ambitious generic operation on
data-type. Given a functor, represented as a tagged description, we
build the free monad over this functor.

@@

Let us recall the free monad construction. Given a functor $F$, the
free monad over $F$ is defined by the following data-type:

@@

Being an inductive type, this $\FMFreeMonad$ data-type is itself
defined by a pattern functor. It is given by:

@@

Of course, we must equip the resulting data-types with operations
delivering a monadic interface. As expected, \(\LAM{\x}\DVar{\x}\)
plays the r\^ole of \return, embedding variables into terms. The
\bind\ operation corresponds to \emph{substitution}. We will now
implement it, as a generic function.

@@

\begin{wstructure}
    -> Consequences
        <- We have free monad data-type
            <- Term + variables
        <- We have monad operations
            <- Return / var
            <- Substitution / bind
\end{wstructure}

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
\V{\sigma}$ completes the implementation. To sum up, we have
implemented the free monad construction for an arbitrary tagged
description. This gives the developer the ability, for any data-type,
to extend it with a notion of variable. Then, we have equipped this
structure with the corresponding monadic operation, \bind\ and
\return. Whereas the \return\ is a trivial variable former, the
\bind\ operation corresponds to the substitution of variables by
terms. We have shown how to implement it, for any free monad. This
programming style is reminiscent to the type-indexed data
types~\cite{hinze:generic-haskell}, as implemented in Generic Haskell:
from data-types, we built new data-types and equip them with some
structure.

\begin{wstructure}
<- Deriving new data-structure and functions on them
    <- Computing the Free Monad of a data-type
        <- Derive new data-structure from previous one
            <- It is just code
        /> New data-structure comes with some equipment
    <- Computing new functions on computed data-types
        <- If data comes with structure, we ought to be able to capture it
            <- Induction on Desc
            -> Ability to compute over data
\end{wstructure}

With the free monad construction, we have seen two kinds of generic
operations. Firstly, we have derived a new data-structure from another
one: we make the free monad from its underlying functor. To do so, we
crucially rely on the fact that data-types are nothing but codes. We
are therefore entitled to modify this code and, in this case, extend
it. Extending a data-type might give rise to a more structured object,
as was the case here.  So, secondly, we have equipped this new
data-type with its inherent structure: the \bind\ and
\return\ operations. We have been able to build them as generic
functions.

>>>>

Apart from these changes, we are left with implementing the other
components of finite sets. This consists of $\EnumT$, $\Ze$, $\Su$, as
well as the $\F{switch}$ operator. Note that the actual implementation
of $\EnumU$ does not influence our implementation of $\EnumT$: be they
hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects
behave the same. We have witnessed this effect when carrying the
operation in Epigram, moving from an hard-coded presentation to a
self-hosted one. Absolutely no change to the $\EnumT$ objects was
required.

\begin{wstructure}
<- Summary of the operation
    <- The content of the type theory is exactly the same
        <- before == after
    /> type naming scheme condenses
        <- Replace named constructors by codes in the universe of datatypes
    -> Our next step is a similar move (in essence)
        /> Condenses the entire naming scheme of datatypes
\end{wstructure}

@@

If not on the content, this modification had an effect on the
names. The type naming scheme of the type theory has condensed: named
type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
now replaced by codes and their fix-point in the universe of
descriptions. In essence, our next step is similar: we are going to
condense the entire naming scheme of datatypes \emph{in itself}.

@@

\begin{wstructure}
<- Realising our promises
    <- We are going to implement Desc
    /> Desc is itself a datatype
        <- Same situation as EnumU
            <- We want to benefit from generic operations
        -> It ought to be encoded in itself
\end{wstructure}

We shall now fulfil our promises for $\Desc$. We are going to
implement the codes of the universe of descriptions. In and by itself,
the system of codes, $\Desc$, is nothing but a datatype. We are in
the same situation as with $\EnumU$: we ought to be able to describe
the codes of $\Desc$ in $\Desc$ itself. Hence, this code would be a
first-class citizen, born with the standard, generic equipment of
datatypes.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the datatypes we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's constructors:
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
that the tagged notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are
defining.

The case of $\DUnit$ is trivial: we immediately reach the end of the
datatype. $\DUnit$ has no payload. The case of $\DIndx{\!}$ comes
naturally: it only takes a recursive argument -- a description. The
case of $\DSigma{\!}{\!}$ is problematic. Recall the specification of
$\DSigma{\!}{\!}$:

@@

Equipped with $\DHindx{\!}{\!}$, we can describe the extended code of
our universe of datatypes, as shown in
Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
remain unchanged, as expected. We successfully describe the
$\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing the $\V{H}$
in $\Set$ with a recursive argument.

At a first glance, we have achieved our goal. We have described the
codes of the universe of description. Taking the fix-point of this
object gives us $\Desc$, up to isomorphism. We have implemented our
code of datatypes as an object levitating inside itself. However,
this levitation operation, just as any magic trick, relies on an
invisible cable. Let us reveal it, hence finishing our implementation.

@@

\begin{wstructure}
<- Generic programming now!
    <- Desc is just data
        -> Can be manipulated
    <- Free induction scheme on Desc
        -> Ability to inspect datatypes
        -> Ability to program on datatypes
\end{wstructure}

@@

This concludes our levitation work. Beyond its pedagogical value, this
exercise has several practical outcomes. First of all, this exercise
reveals that the $\Desc$ universe is just plain data. Just as any
piece of data, it can therefore be inspected and
manipulated. Moreover, it is expressed in the $\Desc$ universe. As a
consequence, it is equipped, for free, with an induction
principle. So, our ability to inspect and program with $\Desc$ is not
restricted to the meta-language: we now have all the necessary
equipment in the target language to \emph{program} over datatypes. In
this setting, \emph{generic programming is just about programming}.

@@

\[\stk{
\F{replace} : \stk{\PITEL{\V{D}}{\Desc}
                   \PITEL{\V{X},\V{Y}}{\Set}\\
                   \PI{\V{xs}}{\descop{\V{D}}{\V{X}}} 
                   \All{\V{D}}{\V{X}}{(\LAM{\_}\V{Y})}{\V{xs}} \To
                   \descop{\V{D}}{\V{Y}}} \\
\F{replace}\: \DUnit\:          \V{X}\: \V{Y}\: \Void\:          \Void          \mapsto 
    \Void                                                                                                         \\
\F{replace}\: (\DSigma{\V{S}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{s}}{\V{xs}}{}\: \V{ys}             \mapsto
    \pair{\V{s}}{\F{replace}\: {\V{D}~\V{s}}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                 \\
\F{replace}\: (\DIndx{\V{D}})\:     \V{X}\: \V{Y}\: \pair{\V{x}}{\V{xs}}{}\: \pair{\V{y}}{\V{ys}}{} \mapsto       \\
\qquad  \pair{\V{y}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{xs}\: \V{ys}}{}                                     \\
\F{replace}\: (\DHindx{\V{H}}{\V{D}})\: \V{X}\: \V{Y}\: \pair{\V{f}}{\V{xs}}{}\: \pair{\V{g}}{\V{ys}}{} \mapsto   \\
\qquad  \pair{\V{g}}{\F{replace}\: \V{D}\: \V{X}\: \V{Y}\: \V{g}\: \V{ys}}{}
}\]

Filling the hole in $\F{cata}$ with \(\F{replace}\: \V{D}\: (\Mu{\V{D}})\: \V{T}\: \V{xs}\: \V{hs}\) closes the
problem. In the type theory, we have built a generic catamorphism. Any
datatype will now come equipped with this operation, for free.

@@

\begin{wstructure}
<- Deriving generic functions
    <- Taking a Desc and computing a function
        <- Desc comes equipped with an induction principle
        -> Ability to compute more functions from it
            -> More generic functions
    <- Inspecting datatypes
        <- All described byu a Desc code
        -> Ability to explore the code
            <- Desc equipped with an induction principle
            -> Build new objects based on that structure
\end{wstructure}

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a datatype specific
operation. This programming style is reminiscent to polytypic
languages, such as PolyP~\cite{jansson:polyp}.

Moreover, the $\F{replace}$ function demonstrates the benefit of an
approach based on universes. The datatypes living in the universe of
descriptions, we are able to \emph{inspect} them. As shown by
$\F{replace}$, it is easy to explore these structures, as well as
building new ones.

@@

In this section, we will turn to a more ambitious generic operation on
datatype. Given a functor, represented as a tagged description, we
build the free monad over this functor.

@@

Let us recall the free monad construction. Given a functor $F$, the
free monad over $F$ is defined by the following datatype:

@@

Being an inductive type, this $\FMFreeMonad$ datatype is itself
defined by a pattern functor. It is given by:

@@

Of course, we must equip the resulting datatypes with operations
delivering a monadic interface. As expected, \(\LAM{\x}\DVar{\x}\)
plays the r\^ole of \return, embedding variables into terms. The
\bind\ operation corresponds to \emph{substitution}. We will now
implement it, as a generic function.

@@

\begin{wstructure}
    -> Consequences
        <- We have free monad datatype
            <- Term + variables
        <- We have monad operations
            <- Return / var
            <- Substitution / bind
\end{wstructure}

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
\V{\sigma}$ completes the implementation. To sum up, we have
implemented the free monad construction for an arbitrary tagged
description. This gives the developer the ability, for any datatype,
to extend it with a notion of variable. Then, we have equipped this
structure with the corresponding monadic operation, \bind\ and
\return. Whereas the \return\ is a trivial variable former, the
\bind\ operation corresponds to the substitution of variables by
terms. We have shown how to implement it, for any free monad. This
programming style is reminiscent to type-indexed
datatypes~\cite{hinze:generic-haskell}, as implemented in Generic
Haskell: from datatypes, we built new datatypes and equip them with
their structure.

\begin{wstructure}
<- Deriving new data-structure and functions on them
    <- Computing the Free Monad of a datatype
        <- Derive new data-structure from previous one
            <- It is just code
        /> New data-structure comes with some equipment
    <- Computing new functions on computed datatypes
        <- If data comes with structure, we ought to be able to capture it
            <- Induction on Desc
            -> Ability to compute over data
\end{wstructure}

%% Candidate for removal:
With the free monad construction, we have seen two kinds of generic
operations. Firstly, we have derived a new data-structure from another
one: we make the free monad from its underlying functor. To do so, we
crucially rely on the fact that datatypes are nothing but codes. We
are therefore entitled to modify this code and, in this case, extend
it. Extending a datatype might give rise to a more structured object,
as was the case here.  So, secondly, we have equipped this new
datatype with its inherent structure: the \bind\ and
\return\ operations. We have been able to build them as generic
functions.
